Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-31 code/plots.Rmd
added escapement appendix

>>>>

<!-- Basic setup -->
```{r setup}
library(tidyverse)
library(patchwork)
library(RColorBrewer)
```

<!-- Data sources -->
```{r}
# for best const mortality and best const escapement
msy_optimizer <- function(df) {
  df |> 
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
  filter(mean_reward == max(mean_reward))
}
escapement_optimizer <- function(df) {
  df |> 
  group_by(escapement, rep) |>
  filter(t == max(t)) |> 
  group_by(escapement) |>
  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
  filter(mean_reward == max(mean_reward))
}

# Collect data (outsource to separate R script later TBD)
msy_df <- read_csv("../data/msy.csv.xz")
best_action <- msy_optimizer(msy_df)$action
escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
best_e <- escapement_optimizer(escapement_df)$escapement
ppo_df <- read_csv("../data/PPO200.csv.xz")

opt_escapement <- escapement_df |> filter(escapement == best_e)

# set up for plots:

msy_sim <- msy_df |>
    filter(action == best_action) |> 
    group_by(rep) |>
    pivot_longer(c("X", "Y", "Z"),
                names_to = "species", values_to = "abundance") |>
    mutate(abundance = abundance + 1, policy = "msy") # natural units, policy name

actions <- unique(msy_df$action)
i <- which.min(abs(actions - best_action* 0.8))
tac_sim <- msy_df |> 
    filter(action == actions[[i]]) |>
    pivot_longer(c("X", "Y", "Z"),
        names_to = "species", values_to = "abundance") |>
    mutate(abundance = abundance + 1, policy = "tac")

ppo_sim <- ppo_df |> 
    pivot_longer(c("X", "Y", "Z"),
        names_to = "species", values_to = "abundance") |>
    mutate(abundance = abundance + 1, policy = "ppo")

esc_sim_df <- escapement_df |> 
    pivot_longer(c("X", "Y", "Z"),
        names_to = "species", values_to = "abundance") |>
    mutate(abundance = abundance + 1)

msy_tac_ppo <- bind_rows(msy_sim, tac_sim, ppo_sim)
```

```{r}
#| label: tuning
#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point. Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet). 
#| (B) Several constant mortality policies shown for illustrative purposes. Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet." 
#| out.width : "6.5in"
msy_rewards <- msy_df |>
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward),
            sd = sd(reward))

rmsy <- msy_rewards |>
  ggplot(aes(action, mean_reward, 
             ymin = mean_reward - 2*sd,
             ymax=mean_reward+2*sd)) +
  geom_point() + geom_ribbon(alpha=0.4) +
  labs(y = "episode reward", x = "constant mortality") +
  theme(aspect.ratio = 1/1) + 
  geom_vline(aes(xintercept=best_action), color="seagreen3", size=1) + 
  geom_vline(aes(xintercept=0.8*best_action), color="violetred3", size=1) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        aspect.ratio = 1)

# linear function blueprint
linear_f <- function(slope, x) slope*x
# generates linear function with given slope
linear_f_gen <- function(slope) (function(x) slope*x)

# dummy data set to keep ggplot happy
p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))

# p + stat_function(fun = linear_f_gen(0.5)) + xlim(0,2)
lines <- p + stat_function(fun = linear_f_gen(0.009)) + 
  stat_function(fun = linear_f_gen(0.018)) +
  stat_function(fun = linear_f_gen(0.027)) +
  stat_function(fun = linear_f_gen(0.8*best_action), col = "violetred3", size=1) +
  stat_function(fun = linear_f_gen(best_action), col = "seagreen3", size=1) +
  stat_function(fun = linear_f_gen(0.054)) + 
  stat_function(fun = linear_f_gen(0.063)) +
  stat_function(fun = linear_f_gen(0.072)) +
  stat_function(fun = linear_f_gen(0.081)) +
  stat_function(fun = linear_f_gen(0.090)) +
  xlim(0,2) +
  xlab("X") + ylab("harvest") +
  theme_bw() +
  theme(#panel.grid.major = element_blank(), 
        #panel.grid.minor = element_blank(), 
        aspect.ratio = 1)

(rmsy | lines) + plot_annotation(tag_levels = c('A'))
```

<!-- Resuts figures -->

```{r} 
#| label: timeseries 
#| fig.cap: "A sample time series for the three fish populations, with the system being controlled, respectively, by the MSY optimal mortality, by $80$ percent of the MSY optimal mortality, and by DRL." 
#| out.width: "6.5in"
policy_labels <- list("msy" = "MSY Const. Mortality", "tac"= "80% MSY Const. Mortality", "ppo"= "DRL")
policy_labeller <- function(variable,value){
  return(policy_labels[value])
}
msy_tac_ppo |> 
    group_by(rep, policy) |>
    filter(rep == 6) |>
    ggplot(aes(t, abundance, col = species)) + 
        geom_line()  + 
        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
        labs(y = "abundance", x = "time") +
        theme_bw() +
        theme(aspect.ratio = 2/3)
```

```{r} 
#| label: rew_len_fig
#| fig.cap: "Bar charts for the episode lengths (top panel) and episode total harvests (bottom panel) for 50 episodes." 
#| out.width: "6.5in"
p1 <- msy_tac_ppo |> 
    group_by(rep, policy) |>
    filter(t==max(t), species == "X", rep < 50) |>
    ggplot(aes(rep, t)) + 
      geom_col() +
      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
      labs(y = "episode length", x = "repetition") +
      theme_bw() + 
      theme(axis.text.x = element_blank())

p2 <- msy_tac_ppo |> 
    group_by(rep, policy) |>
    filter(t==max(t), species == "X", rep < 50) |>
    ggplot(aes(rep, reward)) + 
      geom_col() +
      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
      labs(y = "episode harvest", x = "repetition") +
      theme_bw() + 
      theme(axis.text.x = element_blank())

(p1 / p2) + plot_annotation(tag_levels = 'A')
```

```{r}
#| label: action-X
#| fig.cap: "(A-B) Harvest and escapement scatter data for the DRL policy together with the optimal constant escapement (pink) and constant mortality (green) policies for comparison. 
#| (A) includes $Y$ dependence through color, whereas (B) includes $Z$ dependence similarly.
#| (B-C) escapement (i.e. $X_t - h_t$) scatter data for the DRL policy. 
#| Optimal constant escapement policy shown in pink for comparison."
#| out.width: "6.5in"
opt_escapement_harvest <- function(X) case_when(
  X > best_e ~ X - best_e,
  X <= best_e ~ 0
)

opt_escapement_escapement <- function(X) case_when(
  X <= best_e ~ X,
  X > best_e ~ best_e
)

msy_harvest <- function(X) best_action*X
tac_harvest <- function(X) 0.8*msy_harvest(X)

py <- ppo_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
  ggplot(aes(X, harvest, color = Y)) +
  geom_point() +
  theme_bw() + 
  theme(aspect.ratio = 1/1) +
  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)

pz <- ppo_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action*X) |>
  ggplot(aes(X, harvest, color = Z)) +
  geom_point() +
  theme_bw() + 
  theme(aspect.ratio = 1/1) +
  stat_function(fun = opt_escapement_harvest, col = 'deeppink', size = 0.7) + 
  stat_function(fun = msy_harvest, col = 'seagreen2', size = 0.7)

escxy <- ppo_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
  ggplot(aes(X, escapement, color = Y)) +
  geom_point() +
  theme_bw() + 
  theme(aspect.ratio = 1/1) +
  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)

escxz <- ppo_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X) |>
  ggplot(aes(X, escapement, color = Z)) +
  geom_point() +
  theme_bw() + 
  theme(aspect.ratio = 1/1) +
  stat_function(fun = opt_escapement_escapement, col = 'deeppink', size = 0.7)

((py | pz)/ (escxy | escxz)) + plot_annotation(tag_levels = c('A'))
```
```{r}
#| label: state-space
#| fig.cap: "Policy heatmaps. Each policy, say, $\\pi$, was evaluated at points $(x,y,z)$ visited during any of the 50 evaluation rounds. We projected this data down to the $X$-$Y$ plane---i.e. we plotted the heatmap $\\pi(x,y,z)$ as a function of $x$ and $y$."
#| out.width: "6.5in"
i_tac <- which.min(abs(actions - best_action * 0.8))

# color pallette limits
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
sc <- scale_colour_gradientn(colours = myPalette(100), limits=c(0,0.25))

pmsy <- msy_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
  filter(action == best_action) |>
  ggplot(aes(X, Y, color = harvest)) +
  geom_point() + sc +
  xlim(0,1) +
  ylim(0,0.85) +
  theme_bw() + 
  ggtitle("Opt. Const. Mortality")

ptac <- msy_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
  filter(action == actions[[i_tac]]) |>
  ggplot(aes(X, Y, color = harvest)) +
  geom_point() + sc +
  xlim(0,1) +
  ylim(0,0.85) +
  theme_bw() + 
  ggtitle("80% Opt. Const. Mortality")

pppo <- ppo_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
  ggplot(aes(X, Y, color = harvest)) +
  geom_point() + sc +
  xlim(0,1) +
  ylim(0,0.85) +
  theme_bw() + 
  ggtitle("DRL")

pesc <- opt_escapement |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, 
        harvest = case_when(
          X - escapement > 0 ~ X - escapement,
          X - escapement <= 0 ~  0)
         ) |> 
  ggplot(aes(X, Y, color = harvest)) +
  geom_point() + sc +
  xlim(0,1) +
  ylim(0,0.85) +
  theme_bw() + 
  ggtitle("Opt. Const. Escapement")

(
  (pmsy | ptac) / (pppo | pesc) +
  plot_layout(guides = "collect") & theme(legend.position = 'right')
) + 
  plot_annotation(tag_levels = c('A'))
```

<!-- Appendices -->

```{r} 
#| label: timeseries_msy
#| fig.cap: "Nine time series of the system controlled by the MSY constant mortality policy." 
#| out.width: "6.5in"
msy_tac_ppo |> 
    group_by(rep) |>
    filter(rep < 9, policy == "msy") |>
    ggplot(aes(t, abundance, col = species)) + 
        geom_line()  + 
        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
        facet_wrap(~factor(rep)) +
        labs(y = "abundance", x = "time") +
        theme_bw() + 
        theme(aspect.ratio = 0.5)
```

```{r} 
#| label: timeseries_tac
#| fig.cap: "Nine time series of the system controlled by the policy using 80 percent of the MSY constant mortality." 
#| out.width: "6.5in"
msy_tac_ppo |> 
    group_by(rep) |>
    filter(rep < 9, policy == "tac") |>
    ggplot(aes(t, abundance, col = species)) + 
        geom_line()  + 
        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
        facet_wrap(~factor(rep)) +
        labs(y = "abundance", x = "time") +
        theme_bw() + 
        theme(aspect.ratio = 0.5)
```

```{r} 
#| label: timeseries_ppo
#| fig.cap: "Nine time series of the system controlled by the trained DRL agent." 
#| out.width: "6.5in"
msy_tac_ppo |> 
    group_by(rep) |>
    filter(rep < 9, policy == "ppo") |>
    ggplot(aes(t, abundance, col = species)) + 
        geom_line()  + 
        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
        facet_wrap(~factor(rep)) +
        labs(y = "abundance", x = "time") +
        theme_bw() + 
        theme(aspect.ratio = 0.5)
```

<!-- Old code blocks: -->

### 2023-01-31 manuscript/manuscript.Rmd
added escapement appendix

```{r}
msy_rewards <- msy_df |>
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward),
            sd = sd(reward))

esc_rewards <- escapement_df |>
  group_by(escapement, rep) |>
  filter(t == max(t)) |> 
  group_by(escapement) |>
  summarise(mean_reward = mean(reward),
            sd = sd(reward))

rmsy <- msy_rewards |>
  ggplot(aes(action, mean_reward, 
             ymin = mean_reward - 2*sd,
             ymax=mean_reward+2*sd)) +
  geom_point() + geom_ribbon(alpha=0.4) +
  labs(y = "episode reward", x = "constant mortality") +
  theme(aspect.ratio = 1/1) + 
  geom_vline(aes(xintercept=0.0360), color="red") + 
  geom_vline(aes(xintercept=0.0288), color="green")

resc <- esc_rewards |>
  ggplot(aes(esc_level, mean_reward, 
             ymin = mean_reward - 2*sd,
             ymax=mean_reward+2*sd)) +
  geom_point() + geom_ribbon(alpha=0.4) +
  labs(y = "episode reward", x = "constant escapement") +
  theme(aspect.ratio = 1/1)

# (rmsy | resc)
```

@@

>>>>

@@

## Optimal constant escapement policy evaluation

Here we reproduce the plots from the main text for the optimal constant escapement policy to illustrate the similarity to the DRL-derived policy.

Fig. \ref{fig:esc-tuning-eval} illustrates the tuning process of finding the optimal constant escapment policy (left panel, compare Fig. \ref{fig:tuning}).
The same figure shows, in the middle panel, the reward histogram for the optimal constant escapement policy (compare Fig. \ref{fig:rew_len_fig}).
Finally, the right panel displays the same histogram for the DRL policy (reproduced from Fig. \ref{fig:rew_len_fig}).
We may see that the two policies lead to essentially the same reward distribution.

```{r}
#| label: esc-tuning-eval
#| fig.cap: "(Left panel) Tuning results for constant escapement policies: mean episode reward scatter data (averaged over 50 episodes) together with two standard deviations shaded in grey. 
#| The optimal escapement level (the one producing maximum mean harvest) is $X_{\text{esc.}}=0.62$.
#| (Middle and right panels) Reward histograms for the optimal constant escapement policy (middle) and for the DRL policy (right).
#| For both DRL and optimal constant escapement, no early episode ends are observed (i.e. all episodes last for 200 time steps)."
#| out.width: "6.5in"
#| fig.align: "center"
esc_rewards <- escapement_df |>
  group_by(escapement, rep) |>
  filter(t == max(t)) |> 
  group_by(escapement) |>
  summarise(mean_reward = mean(reward),
            sd = sd(reward))

resc <- esc_rewards |>
  ggplot(aes(escapement, mean_reward, 
             ymin = mean_reward - 2*sd,
             ymax=mean_reward+2*sd)) +
  geom_point() + geom_ribbon(alpha=0.4) +
  labs(y = "episode reward", x = "constant escapement") +
  theme_bw() +
  theme(aspect.ratio = 1/1)

esc_len <- esc_sim_df |>
  group_by(rep) |>
  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
  ggplot(aes(rep, t)) + 
      geom_col() +
      labs(y = "episode length", x = "repetition") +
      theme_bw() + 
      theme(axis.text.x = element_blank())

esc_rew <- esc_sim_df |>
  group_by(rep) |>
  filter(t==max(t), species == "X", escapement == best_e, rep < 50) |>
  ggplot(aes(rep, reward)) + 
      geom_col() +
      labs(y = "episode harvest", x = "repetition", title = "Escapement") +
      ylim(0,12.8) +
      #ggtitle("Escapement") + 
      theme_bw() + 
      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))

drl_rew <- msy_tac_ppo |> 
    group_by(rep, policy) |>
    filter(t==max(t), species == "X", policy == "ppo", rep < 50) |>
    ggplot(aes(rep, reward)) + 
      geom_col() +
      labs(y = "episode harvest", x = "repetition", title = "DRL") +
      ylim(0,12.8) +
      #ggtitle("DRL") + 
      theme_bw() + 
      theme(axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5))

(resc | esc_rew | drl_rew)
```

Fig. \ref{fig:timeseries_esc} displays 9 time-series for episodes controlled with the optimal constant escapement policy.
Again, this behaviour is essentially indistinguishable from the behavior of the system controlled by the DRL policy (see Fig. \ref{fig:timeseries_ppo}).

This series of figures gives evidence that, in fact, the DRL agent we have trained has obtained a policy which, effectively, acts as an optimal constant escapement policy.
Although in Fig. \ref{fig:action-X} we found some deviations from the constant escapement case (e.g. DRL was more conservative when fishing high populations), these do not seem to lead to a different dynamics.

```{r} 
#| label: timeseries_esc
#| fig.cap: "Nine time series of the system controlled by the optimal constant escapement policy." 
#| out.width: "6.5in"
esc_sim_df |> 
    group_by(rep) |>
    filter(rep < 9, escapement == best_e) |>
    ggplot(aes(t, abundance, col = species)) + 
        geom_line()  + 
        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
        facet_wrap(~factor(rep)) +
        labs(y = "abundance", x = "time") +
        theme_bw() + 
        theme(aspect.ratio = 0.5)
```
# References
